---
title: "Tables in R"
output:
  github_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this walkthrough, we'll learn the following key concepts and R functions:

- Using tables to estimate relatively simple probabilities (including conditional and joint probabilities)  
- Making tables with `xtabs`  
- Modifying tables with `prop.table` and `addmargins`  
- piping (`%>%`) as a way of chaining together computations      

## The data

Before you get started, download [aclfest.csv](http://jgscott.github.io/teaching/data/aclfest.csv), which contains data on some bands that played at several major U.S. music festivals (including our our ACL Festival here in Austin).  

## Getting started

We'll first load the [tidyverse library](https://www.tidyverse.org/). 

```{r}
library(tidyverse)
```

If you see a similar set of messages to what's shown above after the `##` symbols, you're good to go!

But if you haven't installed the `tidyverse` library, executing this command will give you an error.  To avoid the error, you'll first need to install `tidyverse` using the `Install` button under the Packages tab, which is typically located in the bottom-right panel of RStudio's four-panel layout.  Remember, an R library is like an app on your phone: you only have to install it once, but you have to load it each time you want to use it.  On a phone, you load an app by clicking on its icon; in R, you load a library using the `library` command.

Then use the `Import Dataset` button to read in the data you downloaded, which you can find under the `Environment` tab in the top-right panel.  Follow the prompts to import `aclfest.csv`.

For a reminder on how to accomplish these two key steps (loading a library, importing a data set), see the previous walkthrough on [Getting Started in R]().  

```{r, echo=FALSE, message=FALSE}
aclfest = read.csv('../../data/aclfest.csv')
```

If you're imported the data correctly, you can run the `head` function to get the first six lines of the file.  You should see the following result:

```{r}
head(aclfest)
```

Each entry is either a 1 or a 0, meaning "yes" and "no", respectively.  So, for example, on the 6th line we see an entry of 1 for Radiohead under the `lollapalooza` column, which means that Radiohead played at Lollapalooza that year.  

Let's make a few simple tables with this data.  This will allow us to estimate probabilities like:  

1. How likely is a band in this sample to play Lollapalooza?  
2. How likely is a band to play ACL Fest, given that they played Lollapalooza?  
3. How like is a band to play either ACL _or_ Lollapalooza?  



## Simple probabilities from tables


Let's address question 1: what is P(played Lollapalooza) for a randomly selected band in this sample?  To answer this, we'll use R's `xtabs` function to tabulate (i.e. count) the data according to whether a band played Lollapalooza (1) or not (0).  

```{r}
xtabs(~lollapalooza, data=aclfest)
```

(You might be curious about the little tilde (`~`) symbol in front of `lollapalooza`.  Roughly speaking, `~` means "by"---as in, "cross-tabulate BY the `lollapalooza` variable.")

OK, back to the table.  Remember that 1 means yes and 0 means no.  So of the 1238 bands (800 + 438) in this sample, 438 of them played Lollapalooza.  We can now just use R as a calculator to get this proportion:

```{r}
438/(800 + 438)
```

### Using `prop.table`

Simple enough, right?  But we can also get R to turn those counts into proportions for us, using the `prop.table` function.  In general, the more work we can get R to do for us, the better!

One way to do this is as follows.  First we'll make the same table as before, except now we'll save the result in an "object" whose name we get to choose.  We'll just call it `t1` here, although you could call it something more imaginative, like `my_great_lollapalooza_table`, if you wanted to.  

```{r}
t1 = xtabs(~lollapalooza, data=aclfest)
```

Notice that nothing gets printed to the screen when you execute this command.  But if you ask R what `t1` is, it will show you the same table as before:
```{r}
t1
```

OK, so why did we bother to store this table in something called `t1`?  Well, remember one of the core ideas in programming: chaining computations together.    That's exactly what we're doing here: we'll take this `t1` object we've created (the first link our chain) and pass it into the `prop.table` function (the second link in our chain).  This function turns a table of counts (like `t1`) into a table of proportions, like this:

```{r}
prop.table(t1)
```

### Using pipes

The above way of doing things---using `xtabs` creating an "intermediate" object called `t1` and then passing `t1` into the `prop.table` function---works just fine.  Lots of people write R code this way.

But it turns out there's a nicer way to accomplish the same task, using a "pipe" (`%>%`).  Here's how it works:
```{r}
xtabs(~lollapalooza, data=aclfest) %>%
  prop.table
```

This code block says: "make a table of counts of the `lollapalooza` variable in the `aclfest` data set, and pipe (`%>%`) the resulting table into `prop.table` to turn the table of counts into a table of proportions."

The result is exactly the same as before; using the pipe is totally optional.  But using pipes tends to make your code easier to easier to write, easier to read, and easier to modify.  For a simple calculation like this involving only two steps, the difference is minimal.  But for the more complex kinds of calculations we'll see later in the course, the difference can be substantial.  I strongly recommend learning to write R code using pipes.


## Joint and conditional probabilities from tables


Let turn to another question: what is P(played ACL | played Lollapalooza)?

To answer this using the data at hand, we'll tabulate the bands by two variables: whether they played at ACL, and whether they played at Lollapalooza.

```{r}
xtabs(~acl + lollapalooza, data=aclfest)
```

These numbers represent joint frequencies:  

- 719 bands played at neither ACL nor Lollapalooza.  
- 81 bands played at ACL but not Lollapalooza.
- 361 bands played at Lollapalooza but not ACL.  
- 77 bands played at both ACL and Lollapalooza.  

To we have to remember the rule for conditional probabilities:

$$
P(A \mid B) = \frac{{P(A,B)}}{P(B)}
$$

The 
