---
output:
  md_document:
    includes:
      in_header: ../header.txt
---

### Boxplots, scatter plots, and lattice plots

In this walk-through, you'll learn how to visualize relationhip of a single quantitative variable.  You will also learn how to change some of the default graphics settings in R plots.

Data files:  
* [ut2000.csv](ut2000.csv): data on SAT scores and graduating GPA for every student who entered the University of Texas at Austin in the fall of 2000 and went on to graduate within 6 years.

First, in order to complete this analysis we need to load in the `mosaic` library from within RStudio. An R library (or package) is bundle of commands that provides additional functionality, beyond what comes with the basic R installation. There are literally thousands of packages available for R, ranging from the simple to the very sophisticated. The mosaic package was written specifically for use in statistics classrooms. We will use it along with a handful of other packages this semester, so you’ll need to learn how to install them. The first minute of [this video](https://www.youtube.com/watch?v=u1r5XTqrCTQ) gives a walk-through, but it's quite simple: under the Packages tab, there's a button for install.  Click it, type in the package name (it should auto-complete), click install.

Once the installation is done, you'll be ready to use the library.  Load it in with the `library` command:
```{r}
library(mosaic)
```
There will be a lot of information returned to the console when you load the library, but you only have to worry if you see an actual error, like this:
```
Error in library(mosaic) : there is no package called ‘mosaic’
```
This means you haven't installed the package!  You'll only have to do the installation once, but you will have to load the mosaic library (as above) at the beginning of each new R session if you intend to use it.

Next, download the ut2000.csv file and read it in.
```{r}
ut2000 = read.csv('ut2000.csv')
summary(ut2000)
```

School (which of the 10 different undergraduate schools at UT the student graduated from) is a natural grouping variable here.  Let's start by computing the mean SAT math score stratified by school.
```{r}
mean(SAT.Q ~ School, data=ut2000)
```

This gives you an idea of between-group variation.  To see both the between-group and within-group variation, you'll need a boxplot then a boxplot of SAT math scores (SAT.Q)
```{r, fig.width=9}
bwplot(SAT.Q ~ School, data=ut2000)
```

The `bwplot` command is available because we loaded the `mosaic` package. You'll notice the aesthetics are a little nicer than the basic R command `boxplot`.  You may also notice the names of the colleges along the x axis running together, which you can fix by clicking "Zoom" in the Plots tab and manually resizing the window.

Next, let's look at a scatterplot of graduating GPA versus SAT math scores for all students.
```{r}
plot(GPA ~ SAT.Q, data=ut2000)
```

If we want to see this relationship separately for each of the 10 colleges, we need what's called a lattice plot.  The `xyplot` command produces one:
```{r}
xyplot(GPA ~ SAT.Q | School, data=ut2000)
```

The vertical bar (|) should be read as "conditional upon" or "stratified by."
