---
fig.width: 6
output:
  md_document:
    includes:
      in_header: ../header.txt
---

### Measuring and visualizing dispersion

In this walk-through, you'll learn how to measure and visualize dispersion of a single quantitative variable.  You will also learn how to change some of the default graphics settings in R plots.

Data files:  
* [citytemps.csv](citytemps.csv): daily average temperatures in San Diego, CA and Rapid City, SD.  

First download the citytemps.csv file and read it in. You can use RStudio's Import Dataset button, or use the read.csv command:
```{r}
citytemps = read.csv('citytemps.csv')
```
If you do this second option, make sure you're working from the directory where the data lives, or put the full path to the file between the quotation marks.

We'll start off by getting R to summarize the variables:

```{r}
summary(citytemps)
```

You can see we have 17 years of temperature data on San Diego and Rapid City.  Let's focus on San Diego. We'll start with a histogram, one of the most basic plots for a single quantitative variable:

```{r}
hist(citytemps$Temp.SanDiego)
```

You already know one way of summarizing the dispersion of this distribution: the standard deviation.
```{r}
mean(citytemps$Temp.SanDiego)
sd(citytemps$Temp.SanDiego)
```


Another measure of dispersion is the coverage interval: that is, an interval covering a specified fraction of the observations.  For example, to get a central 50% coverage interval, we'd need the 25th and 75 percentiles of the distribution.  By definition, 50% of the observations are between these two numbers.  You can get these from the quantile function.
```{r}
quantile(citytemps$Temp.SanDiego)
```

You can also ask for different quantiles by passing a second argument called "probs," like this:
```{r}
quantile(citytemps$Temp.SanDiego, probs=c(0.05, 0.95))
```
In computer science an "argument" is just something we pass to a function.  In this case, `probs=c(0.05, 0.95)` is an optional argument that modifies the default behavior of the quantile function to yield the 5th and 95th percentiles.  We could use these numbers to construct a 90% coverage interval.

### Changing the default look of plots

Next, we'll make a histogram for Rapid City, except we'll change the default title and x-axis label to something a bit more fit for human consumption.
```{r}
hist(citytemps$Temp.RapidCity, main="Average Daily Temperatures in Rapid City, 1995-2011", xlab='Temperature')
```

Notice again that we've used optional arguments (main and xlab) to change the main title and x-axis label, respectively.

Next, let's stack the two histograms on top of each other, to make a multi-frame plot.  We do this with the following code block.  The first line says to make a multi-frame plot (filled in along the rows) with 2 rows and 1 column.  The next two plotting commands then fill in the two frames.

```{r, fig.height=8}
par(mfrow=c(2,1))
hist(citytemps$Temp.SanDiego)
hist(citytemps$Temp.RapidCity)
```

This won't do at all: notice that the axes and bin sizes differ between the two plots.  This makes it hard to compare the two distributions at a glance.  We need to align these two plots to have the same axes and bins.  Just as we did above, we'll do this by passing additional arguments to the `hist` function.

First, we must define a set of breakpoints for the histogram grams.  We'll do this with the `seq` (which stands for sequence) command:
```{r}
mybreaks = seq(-20, 92, by=2)
```
This says to make a sequence running from -20 to 92 degrees in increments of 2.  You'll see the whole sequence if you type `mybreaks` directly into the console"
```{r}
mybreaks
```

Now we can make the histograms using these custom bins.  We'll also change the x and y axes using the xlim and ylim arguments:
```{r, fig.height=8}
par(mfrow=c(2,1))
hist(citytemps$Temp.SanDiego, breaks=mybreaks, xlim=c(-20,100), ylim=c(0, 760))
hist(citytemps$Temp.RapidCity, breaks=mybreaks, xlim=c(-20,100), ylim=c(0, 760))
```

Now the bins and axes are comparable, making the distributions themselves much easier to compare.

### Optional

Below, I've included a set of commands that produce a much prettier plot.  Feel free to go line by line and modify these commands to see what behavior they produce.  You can learn a lot about R this way:

```{r}
mybreaks = seq(-20, 92, by=2)
par(mfrow=c(1,1), mar=c(3,0,1,3), mgp=c(2,1,0))
hist(citytemps$Temp.SanDiego, breaks=mybreaks, xlab="Average Daily Temperature (F)", main="", border="darkgrey", col="grey", axes=FALSE, ylim=c(0, 760))
hist(citytemps$Temp.RapidCity,breaks=mybreaks,add=TRUE, border=rgb(0,100,0,100,maxColorValue=255), col= rgb(0,100,0,50,maxColorValue=255))
axis(4,at=seq(0,700,by=100), las=1,tick=FALSE)
axis(1,pos=0)
text(55, 770, "San Diego, CA", pos=4, font=2)
text(30, 260, "Rapid City, SD", pos=4, font=2)
```

